Algorithms for Supervised Dimensionality Reduction with Random Projection-Based Ensemble Learning

### 2021-02-10 main.tex

@@

\maketitle
\section{Introduction}

\subsection{Terminology and Basic Concepts}
We will be discussing high-dimensional regression and classification methods. These are supervised methods, by which we mean that we are presented with labelled training data 
\[ \mathcal{T}_n = \{(x_1, y_1), (x_2, y_2), \dots, (x_n, y_n) \ | \ x_i \in \mathcal{X}, y_i \in \mathcal{Y}\}. \]
We refer to $\mathcal{X}$ as the 'input space' or sometimes 'observation space', and will typically assume it is a subset of real Euclidean space $\mathcal{X} \subset \mathbb{R}^p$. We refer to $\mathcal{Y}$ as the 'outcome space' or 'label space'. In the context of regression it may be some subset of the real line, whereas in classification it is a discrete, finite space. We will mostly consider classification, and assume for simplicity that $\mathcal{Y} = \{0,1\}$. The elements of $\mathcal{T}_n$ can be considered as observations of a random variable $(X,Y)$ from a joint probability distribution $P_{X \times Y}$, although in some circumstances it will be more convenient to think of the values $x_i$ as fixed. We define a classifier as a measurable function $C: \mathcal{X} \rightarrow \mathcal{Y}$, and denote by $\mathcal{C}_{\mathcal{X},\mathcal{Y}}$ the space of all such functions (note that this is identical for regression apart from the set $\mathcal{Y}$ and the measure used). A learning procedure is a function $L : \mathcal(\mathcal{X} \times \mathcal{Y})^n \rightarrow \mathcal{C}_{\mathcal{X},\mathcal{Y}}$, i.e. a procedure that produces a classifier from a set of training data. When the learning procedure used is clear (or irrelevant) we will write $C_{\mathcal{T}_n} = L(\mathcal{T}_n)$ to refer to a classifier trained on the given training data. Common learning procedures for classification include linear and quadratic discriminant analysis, $k$-nearest neighbours, logistic regression, support vector machines and random forests.

We can measure the performance of a classifier via its risk, given by 
\[R(C) = \mathbb{E}_{(X,Y)}[\mathbbm{1}_{\{C(X) \neq Y\}}] \quad \Big( = \int_{\mathcal{X}\times \mathcal{Y}} \mathbbm{1}_{\{C(X) \neq Y\}} dP_{X\times Y}\Big),\]
where we write in terms of the expectation of $\mathbbm{1}_{\{X(X)\neq Y\}}$ rather than the probability of $\{C(X) \neq Y\}$ (they are equivalent) to emphasise the close analogy with regression risk, in which the discrete distance $\mathbbm{1}_{\{C(X) \neq Y \}}$ is replaced by the $L_2$ distance $||C(X)-Y||_2^2$. In practice the quantity $R(C)$ is often intractable, and we have to estimate it with a random variable $\hat{R}$. Choices for $\hat{R}$ include the training error $\hat{R}(C) := \frac{1}{n}\sum_{(x,y) \in \mathcal{T}_n} \mathbbm{1}_{\{C(x) \neq y\}}$, or the test error 
\[\hat{R}(C) = \frac{1}{m} \sum_{(x,y) \in \mathcal{T}_{m}} \mathbbm{1}_{\{C(x) \neq y\}},\] 
where unseen observation-outcome pairs come from a second dataset $\mathcal{T}_{m} = \{(x_i, y_i) \ | \ 1 \leq i \leq m \}$.

All of the methods mentioned above suffer from a phenomenon known as the curse of dimensionality when $p$ is large. Loosely put, the curse of dimensionality refers to the observation that the volume of a region of space (such as a hypercube) increases exponentially with $p$, necessitating an exponential increase in sample size to maintain the same density of observations and so predictive performance. This is true even when dimensionality $p$ and sample size $n$ increase together at the same rate with $p<n$ (for a thorough treatment see \citet{Wainwright2019}), but is particularly exacerbated when $p \gg n$. In this regime overfitting becomes very difficult to manage. In this situation it is common to make some assumption of lower dimensional structure. Common practices for imposing lower dimensional structure include sample clustering and variable selection. Both can be implemented in a variety of ways and are liable to fail in certain situations, as we shall discuss in the next section.

@@

\subsection{Sufficient  and Central Subspaces}
In order to address the issues addressed in the previous section, we seek a notion of dimensionality reduction that is \textit{a)} informed by labels as well as observations and \textit{b)} rotationally invariant. To this end we consider subspace selection as a generalisation of variable selection. Intuitively, we seek a low-dimensional subspace $V \subset \mathcal{X}$ such that almost all information pertaining to label value is encapsulated by an observation's projection onto $V$. If $V$ can be expressed as the span of canonical basis (i.e. axis-aligned) vectors, then this is equivalent to variable selection.

@@

We may immediately see that all joint distributions $(X,Y)$ admit a sufficient subspace (namely via the identity map $A= I_p$). Furthermore, any superspace $W \supseteq V$ of a sufficient subspace $V$ is also sufficient. While it is not necessarily true that the intersection $V_1 \cap V_2$ of two sufficient subspaces $V_1,V_2$ is itself sufficient, \citet[Lemma 2]{Cook1996} laid out conditions under which this is the case which are broad enough for our purposes. Under such circumstances it makes sense to discuss the central subspace 
\begin{equation*}
    C_{Y|X} := \cap_{V: V\text{ sufficient}}V.
\end{equation*}
A variety of methods for estimation of the central subspace have been proposed. Several have proceeded by maximising an estimate of the dependence between $AX$ and $Y$, where the dependence is measured by log-loss mutual information \citep{torkkola_feature_2003, suzuki_approximating_2008}, squared-loss mutual information \citep{suzuki_sufficient_2012, yamada_sufficient_2011}, or distance correlation \citep{vepakomma_supervised_2018}. Related to the notion of the central subspace is that of the central mean subspace,introduced in \citet{cook_dimension_2002} along with a discussion of methods for its estimation. In the case of binary classification, which we will consider, the central subspace and central mean subspace are equivalent. Therefore, work geared towards estimation of the central mean subspace \citep[see for example][]{ma_estimation_2014} is also within our purview.\\

@@

\subsection{False discovery and power for dimension reduction}

@@

   Classify $x$ with label 0\;
  }
 \caption{Random Projection Ensemble Classifier\label{alg:rpclass}}
\end{algorithm} 
~\\
The random projections in Algorithm \ref{alg:rpclass} may be chosen according to Haar measure, to be axis-aligned, or to be sparse in some sense. In some situations it is acceptable not to insist they be true projections at all, and simply choose a random matrix of the correct side. The parameter $\alpha$ intuitively may be chosen to be $1/2$, but this is not always actually optimal: discussion of how to choose $\alpha$ can be found in \citet{Cannings2017}.

\section{Subspace estimation based on the random-projection ensemble classifier}
In \citet{Cannings2017} the random-projection ensemble classifier was shown to have excellent predictive performance. However, it does not immediately admit a procedure for variable or subspace selection. While each classifier is based on a $d$-dimensional representation of the observations, the ensembling procedure means that the final set of features used to make a prediction will almost certainly span the entire space $\mathbb{R}^p$ if $B_1$ and $d$ are large enough (if $B_1d \geq p$).

@@

@@

    \begin{subfigure}[t]{0.45\textwidth}
        \centering
        \includegraphics[height=2.2in]{complexfirst.png}
        \caption{}
    \end{subfigure}
    ~
    \begin{subfigure}[t]{0.45\textwidth}
        \centering
        \includegraphics[height=2.2in]{complexsecond.png}
        \caption{}
    \end{subfigure}
\caption{}
\end{figure*}
\section{Real-World Examples}
\subsection{Epilepsy Dataset}
\section{Generalised formulation?}
If we made good progress on the techniques that we've got at the moment, I think there would be scope for looking at a range of generalisations. These can be grouped into alterations of the selection method for choosing random projections, and alterations of the distance measure used to obtain low-dimensional representations. Consider the following (very general) setup, where we have a few objects:
\begin{itemize}
    \item A training set $\mathcal{T}_n$.
    \item A collection of projections $\mathcal{A} := \{ A_b \in \mathbb{R}^{d\times p} : b \in \{ 1,\dots,B\}$ (by projections we mean that for all $b$, $A_bA_b^T = I_d$).
    \item A base classifier, i.e. a map $C : (\mathcal{T}_n, A) \mapsto C^A_{\mathcal{T}_n}$ where $C^A_{\mathcal{T}_n}$ is a function $C^A_{\mathcal{T}_n}: \mathcal{X} \rightarrow \mathcal{Y}$.
    \item A selection procedure, by which we mean a map $s : (\mathcal{T}_n, \mathcal{A}, b, C) \mapsto s_b$ where $s_b:\mathcal{X}^2 \rightarrow \{0,1\}$ is a function detailing which pairs of observations to select that projection for. 
    \item A positive definite kernel $k: \mathbb{R}^d \times \mathbb{R}^d \rightarrow \mathbb{R}$.
\end{itemize}

Now we define a learned distance

\[K^\mathcal{A}_{\mathcal{T}_n}(x,x') = \sum_{b: s_b(x,x') = 1} k(A_bx, A_bx').\]

>>>>

\usepackage[acronym, toc]{glossaries-extra}

\setabbreviationstyle[acronym]{long-short}

\glssetcategoryattribute{acronym}{nohyperfirst}{true}

\renewcommand*{\glsdonohyperlink}[2]{%
 {\glsxtrprotectlinks \glsdohypertarget{#1}{#2}}}

\makeglossaries

\newacronym{sdr}{SDR}{Supervised Dimensionality Reduction}

@@

\maketitle

\begin{abstract}
We propose a class of algorithms for \glsxtrlong{sdr} based on ensembles of random projections coupled with a low-dimensional classifier. Our core method is agnostic to the learning procedure employed, and can produce informative representations for a variety of high-dimensional data types.
\end{abstract}

\section{Introduction}
\gls{sdr} entails producing low-dimensional representations of high-dimensional objects, which preserve information with respect to a target variable. Applications of \gls{sdr} are found in data compression, visualisation, and as a pre-training step for further predictive models. We propose a class of algorithms for \gls{sdr} based on ensembles of random projections coupled with a low-dimensional classifier. Our core method is agnostic to the learning procedure employed, and can produce informative representations for a variety of high-dimensional data types.

\subsection{Terminology and Basic Concepts}
We will be discussing supervised classification methods. By supervised, we mean that we are presented with labelled training data 
\[ \mathcal{T}_n = \{(x_1, y_1), (x_2, y_2), \dots, (x_n, y_n) \ | \ x_i \in \mathcal{X}, y_i \in \mathcal{Y}\}. \]
We refer to $\mathcal{X}$ as the 'input space' or 'observation space', and will assume it is a subset of real Euclidean space so that $\mathcal{X} \subset \mathbb{R}^p$. We refer to $\mathcal{Y}$ as the 'outcome space' or 'label space', and assume for simplicity that $\mathcal{Y} = \{0,1\}$. The elements of $\mathcal{T}_n$ can be considered as observations of a random variable $(X,Y)$ from a joint probability distribution $P_{X \times Y}$, although in some circumstances it will be more convenient to think of the values $x_i$ as fixed. We define a classifier as a measurable function $C: \mathcal{X} \rightarrow \mathcal{Y}$, and denote by $\mathcal{C}_{\mathcal{X},\mathcal{Y}}$ the space of all such functions (note that this is identical for regression apart from the set $\mathcal{Y}$ and the measure used). A learning procedure is a function $L : \mathcal(\mathcal{X} \times \mathcal{Y})^n \rightarrow \mathcal{C}_{\mathcal{X},\mathcal{Y}}$, i.e. a map from training data sets to classifiers. When the learning procedure used is clear (or irrelevant) we will write $C_{\mathcal{T}_n} = L(\mathcal{T}_n)$ to refer to a classifier trained on the given training data. Common learning procedures for classification include linear and quadratic discriminant analysis, $k$-nearest neighbours, logistic regression, support vector machines and random forests.

We can measure the performance of a classifier via its risk, given by 
\[R(C) = \mathbb{E}_{(X,Y)}[\mathbbm{1}_{\{C(X) \neq Y\}}] \quad \Big( = \int_{\mathcal{X}\times \mathcal{Y}} \mathbbm{1}_{\{C(X) \neq Y\}} dP_{X\times Y}\Big).\]
Note that this definition has a close analogy with regression risk, in which the discrete distance $\mathbbm{1}_{\{C(X) \neq Y \}}$ is replaced by the $L_2$ distance $||C(X)-Y||_2^2$. In practice the quantity $R(C)$ is often intractable, and we have to estimate it with a random variable $\hat{R}$. Choices for $\hat{R}$ include the training error $\hat{R}(C) := \frac{1}{n}\sum_{(x,y) \in \mathcal{T}_n} \mathbbm{1}_{\{C(x) \neq y\}}$, or the test error 
\[\hat{R}(C) = \frac{1}{m} \sum_{(x,y) \in \mathcal{T}_{m}} \mathbbm{1}_{\{C(x) \neq y\}},\] 
where unseen observation-outcome pairs come from a second dataset $\mathcal{T}_{m} = \{(x_i, y_i) \ | \ 1 \leq i \leq m \}$.

All of the methods mentioned above suffer from a phenomenon known as the curse of dimensionality when $p$ is large \citep{mai_review_2013}. Loosely put, the curse of dimensionality refers to the observation that the volume of a region of space (such as a hypercube) increases exponentially with $p$, necessitating an exponential increase in sample size to maintain the same density of observations and so predictive performance. This is true even when dimensionality $p$ and sample size $n$ increase together at the same rate with $p<n$ (for a thorough treatment see \citet{wainwright_high-dimensional_2019}), but is particularly exacerbated when $p \gg n$. In this regime overfitting becomes very difficult to manage. In this situation it is common to make some assumption of lower dimensional structure. Common practices for imposing lower dimensional structure include sample clustering and variable selection. Both can be implemented in a variety of ways and are liable to fail in certain situations, as we shall discuss in the next section.

@@

\subsection{Sufficient  and Central Subspaces}
In order to address the issues established in the previous section, we seek a notion of dimensionality reduction that is \textit{a)} informed by labels as well as observations and \textit{b)} rotationally invariant. To this end we consider subspace selection as a generalisation of variable selection. Intuitively, we seek a low-dimensional subspace $V \subset \mathcal{X}$ such that almost all information pertaining to label value is encapsulated by an observation's projection onto $V$. If $V$ can be expressed as the span of canonical basis (i.e. axis-aligned) vectors, then this is equivalent to variable selection.

@@

We may immediately see that all joint distributions $(X,Y)$ admit a sufficient subspace (namely via the identity map $A= I_p$). Furthermore, any superspace $W \supseteq V$ of a sufficient subspace $V$ is also sufficient. While it is not necessarily true that the intersection $V_1 \cap V_2$ of two sufficient subspaces $V_1,V_2$ is itself sufficient, \citet[Lemma 2]{cook_graphics_1996} laid out conditions under which this is the case which are broad enough for our purposes. Under such circumstances it makes sense to discuss the central subspace 
\begin{equation*}
    C_{Y|X} := \cap_{V: V\text{ sufficient}}V.
\end{equation*}
A variety of methods for estimation of the central subspace have been proposed. Several have proceeded by maximising an estimate of the dependence between $AX$ and $Y$, where the dependence is measured by log-loss mutual information \citep{torkkola_feature_2003, suzuki_approximating_2008}, squared-loss mutual information \citep{suzuki_sufficient_2012, yamada_sufficient_2011}, or distance correlation \citep{vepakomma_supervised_2018}. Related to the notion of the central subspace is that of the central mean subspace,introduced in \citet{cook_dimension_2002} along with a discussion of methods for its estimation. In the case of binary classification, which we will consider, the central subspace and central mean subspace are equivalent. Therefore, work geared towards estimation of the central mean subspace \citep[see for example][]{ma_estimation_2014} is also within our purview.\\

@@

\subsection{False discovery and power for dimension reduction}
\citep{taeb_false_2020}

@@

   Classify $x$ with label 0\;
  }
 \caption{Random Projection Ensemble Classifier\label{alg:rpclass}}
\end{algorithm} 
~\\
The random projections in Algorithm \ref{alg:rpclass} may be chosen according to Haar measure, to be axis-aligned, or to be sparse in some sense. In some situations it is acceptable not to insist they be true projections at all, and simply choose a random matrix of the correct side. The parameter $\alpha$ intuitively may be chosen to be $1/2$, but this is not always actually optimal: discussion of how to choose $\alpha$ can be found in \citet{cannings_random-projection_2017}.

\section{Subspace estimation based on the random-projection ensemble classifier}
In \citet{cannings_random-projection_2017} the random-projection ensemble classifier was shown to have excellent predictive performance. However, it does not immediately admit a procedure for variable or subspace selection. While each classifier is based on a $d$-dimensional representation of the observations, the ensembling procedure means that the final set of features used to make a prediction will almost certainly span the entire space $\mathbb{R}^p$ if $B_1$ and $d$ are large enough (if $B_1d \geq p$).

@@

\citep{paulin_efronstein_2016}

@@

    \begin{subfigure}[t]{0.45\textwidth}
        \centering
        \includegraphics[height=2.2in]{complexfirst.png}
        \caption{}
    \end{subfigure}
    ~
    \begin{subfigure}[t]{0.45\textwidth}
        \centering
        \includegraphics[height=2.2in]{complexsecond.png}
        \caption{}
    \end{subfigure}
\caption{}
\end{figure*}
\section{Real-World Examples}
\subsection{Epilepsy Dataset}

% \section{Generalised formulation?}
% If we made good progress on the techniques that we've got at the moment, I think there would be scope for looking at a range of generalisations. These can be grouped into alterations of the selection method for choosing random projections, and alterations of the distance measure used to obtain low-dimensional representations. Consider the following (very general) setup, where we have a few objects:
% \begin{itemize}
%     \item A training set $\mathcal{T}_n$.
%     \item A collection of projections $\mathcal{A} := \{ A_b \in \mathbb{R}^{d\times p} : b \in \{ 1,\dots,B\}$ (by projections we mean that for all $b$, $A_bA_b^T = I_d$).
%     \item A base classifier, i.e. a map $C : (\mathcal{T}_n, A) \mapsto C^A_{\mathcal{T}_n}$ where $C^A_{\mathcal{T}_n}$ is a function $C^A_{\mathcal{T}_n}: \mathcal{X} \rightarrow \mathcal{Y}$.
%     \item A selection procedure, by which we mean a map $s : (\mathcal{T}_n, \mathcal{A}, b, C) \mapsto s_b$ where $s_b:\mathcal{X}^2 \rightarrow \{0,1\}$ is a function detailing which pairs of observations to select that projection for. 
%     \item A positive definite kernel $k: \mathbb{R}^d \times \mathbb{R}^d \rightarrow \mathbb{R}$.
% \end{itemize}

% Now we define a learned distance

% \[K^\mathcal{A}_{\mathcal{T}_n}(x,x') = \sum_{b: s_b(x,x') = 1} k(A_bx, A_bx').\]
